---
title: "First approach to service availability"
output:
  pdf_document:
    toc: true
    toc_depth: 2
---

This notebook is a first exploration of how SNIS and RBF data collected in DRC through dedicated DHIS2 portals can be used to measure availability of vaccination services. For each of these portals, all available metadata has been extracted using the package DHISExtractr. For the SNIS portal, the following elements have also been extracted:

* The list of indicators mentioning the initials "BCG"
* The list of facilities supposed to be reporting on at least one of these indicators. 
* All available data for these indicators in these facilities for the years 2016-2017. Looking at the data, it turns out there has been a structural change in the level of data availability in 2017, and we thus dropped the (very) few available 2016 data.

For legibility of this notebook, all _housekeeping_ code has been muted.

```{r echo=FALSE , message=FALSE , warning=FALSE , error=FALSE}
library(dplyr)
library(ggplot2)
library(dhisextractr)
library(rgdal)
library(pals)
library(classInt)
library(latticeExtra)
library(grid)

load_env('../../.env')

setwd(work_dir)

data_elements <- read.csv(paste0(snis_data_dir, '/data_elements_metadata.csv'))
org_units_reports <- read.csv(paste0(snis_data_dir, '/org_units_report.csv'))
data_elements_sets <- read.csv(paste0(snis_data_dir, '/data_elements_sets.csv'))
snis_org_units <- read.csv(paste0(snis_data_dir, '/org_units_description.csv'))

data_elements_pdss <- read.csv(paste0(pdss_data_dir, '/data_elements_metadata.csv'))
org_units_reports_pdss <- read.csv(paste0(pdss_data_dir, '/org_units_report.csv'))
data_elements_sets_pdss <- read.csv(paste0(pdss_data_dir, '/data_elements_sets.csv'))

data_elements_bcg <- look_up_data_element_term('Artesunate-Amodiaquine')
data_sets_bcg <- look_up_data_set_term('Artesunate-Amodiaquine')
org_units_bcg <- get_de_reporting_facilities(data_sets_bcg)

data_pdss_full <- read.csv('data/raw/pdss_data.csv')
data_snis_full <- read.csv('data/raw/snis_malar_data.csv')
fac_zones <- read.csv('data/references/snis_fosas_zones.csv')
fac_zones_pdss <- read.csv('data/references/pdss_fosas_zones.csv')

matched_zones <- read.csv('data/references/matched_zones.csv')

year = '2017'
start <- paste0(year, '01')
end <- paste0(year, '12')

data <- data_snis_full[data_snis_full$period >= start & data_snis_full$period <= end, ]
data <- merge(data, fac_zones, by.x = 'org_unit_ID', by.y = 'units')

data_pdss <- data_pdss_full[as.character(data_pdss_full$period) > "2016Q4", ]
data_pdss <- merge(data_pdss, fac_zones_pdss, by.x = 'org_unit_ID', by.y = 'units')
data_pdss <- merge(data_pdss, data_elements_pdss, by.x = 'data_element_ID' , by.y = 'id')


org_units_bcg <- fac_zones[fac_zones$units %in% org_units_bcg,]

shapefile_data <- readOGR(paste0(snis_data_dir, '/map_polygons.shp'))
coordinates <- fortify(shapefile_data, region="org_unit_I")
colnames(coordinates) <- c('long','lat','order','hole','piece','zone','group')
coordinates <- coordinates[coordinates$zone %in% fac_zones$zone,]
```


## Data Completeness

We first want to look at the completeness of the data in the SNIS DHIS reporting. To do this by zone, we consider each facility in a given zone should have reported 12 times in the year 2017. 

```{r message=FALSE, warning=FALSE}
fosa_by_zone <- group_by(org_units_bcg, zone)
n_expected <- summarise(fosa_by_zone, n_expected = length(unique(as.character(units))))

map_expectation <- right_join(coordinates , n_expected, by = 'zone')

ggplot(map_expectation)+
  geom_polygon(aes(x  = long, y = lat , 
                   group=zone, 
                   fill = n_expected)) +
  theme_minimal() +
  coord_map()+ 
  scale_fill_distiller(name = "Expected Reports",
                       palette = "Spectral" , direction = -1)+ 
  theme(legend.position = "bottom")
```

We can now compare this data to the number of reports actually obtained by zone on each of the target indicators.

```{r message=FALSE, warning=FALSE}
data_by_zone_period <- group_by(data, zone , period)
n_reporting_period <- summarise(data_by_zone_period, n_reporting_monthly = length(unique(as.character(org_unit_ID))))

n_reporting_zone <- group_by(n_reporting_period, zone)
n_reporting_zone <- summarise(n_reporting_zone, n_reporting = sum(n_reporting_monthly))

n_reporting <- inner_join(n_reporting_zone , n_expected)
n_reporting$reporting_rate <- n_reporting$n_reporting / (12 * n_reporting$n_expected)
n_reporting <- n_reporting[n_reporting$reporting_rate <= 1 , ]


map_reporting_rates <- right_join(coordinates , n_reporting, by = 'zone')
ggplot(map_reporting_rates)+
  geom_polygon(aes(x  = long, y = lat , 
                   group=zone, 
                   fill = reporting_rate)) +
  theme_minimal() +
  coord_map()+ 
  scale_fill_distiller(name = "Reporting Rates",
                       palette = "Spectral" , direction = 1)+ 
  theme(legend.position = "bottom")
```

## BCG stockouts description

### Based on reported stockouts

First, let's look at the reported stockouts.

```{r , warning=FALSE}
data_rupture <- data[(data$name == 'B 6.1 DTC-HepB-Hib doses - jours de rupture de stock' | 
                        data$name == 'F 4.1 DTC-HepB-Hib doses - Jours de rupture de stock') & 
                            data$value < 32 ,]

stockout_by_zone <- group_by(data_rupture , zone)
mean_reported_stockout <- summarize(stockout_by_zone, mean_rupture = mean(value))

map_reported_stockout <- right_join(coordinates , mean_reported_stockout, by = 'zone')
ggplot(map_reported_stockout)+
  geom_polygon(aes(x  = long, y = lat , 
                   group=zone, 
                   fill = mean_rupture)) +
  theme_minimal() +
  coord_map()+ 
  scale_fill_distiller(name = "Mean reported stockout days",
                       palette = "Spectral" , direction = -1)+ 
  theme(legend.position = "bottom")
```

Trying to look at both reporting and ruptures at the same time.

```{r}
# Based on https://cran.r-project.org/web/packages/pals/vignettes/bivariate_choropleths.html

cols <- census.blueyellow ; nbins <- 3

map_bivariate <- merge(map_reported_stockout, map_reporting_rates, by=c('long','lat','order','hole','piece','zone','group'))

color_rupture <- findCols(classIntervals(map_bivariate$mean_rupture, n=nbins, style='fixed',
                                           fixedBreaks=c(0,5, 15,31)))
color_reporting <- findCols(classIntervals(map_bivariate$reporting_rate, n=nbins, style='fixed',
                                           fixedBreaks=c(0,.5,.80,1.1)))

map_bivariate$color_class <- color_reporting + nbins*(color_rupture-1)

plot(mean_rupture ~ reporting_rate, data=map_bivariate,
     col=cols()[map_bivariate$color_class])

#map_bivariate <- map_bivariate[order(map_bivariate$zone),]

#Cmap <- map(list(x = map_bivariate$long, y = map_bivariate$lat, names = map_bivariate$group), plot = FALSE, fill = FALSE)#,
#                        projection = "tetra")
colnames(shapefile_data@data) <- 'name'
Cmap <- shapefile_data

m3 <- mapplot(zone ~ color_class, data = map_bivariate,
              colramp=cols, breaks=seq(from=0.5, by=1, length=nbins*nbins+1),
              xlab="",
              colorkey=FALSE,
              map = map(Cmap),
              scales = list(draw = FALSE))
suppressWarnings(print( m3 ))
# add the color legend
m3leg <- levelplot(matrix(1:(nbins*nbins), nrow=nbins), axes=FALSE, col.regions=cols(),
                   xlab="completeness  -->", ylab="rupture  -->", cuts=8, colorkey=FALSE,
                   scales=list(draw=0))
vp <- viewport(x=.15, y=.25, width=.4, height=.4)
pushViewport(vp)
print(m3leg, newpage=FALSE)
popViewport()

map_reported_stockout_2 <- merge(map_reported_stockout, map_reporting_rates, by=c('long','lat','order','hole','piece','zone','group') , sort = FALSE)
ggplot(map_reported_stockout_2)+
  geom_polygon(aes(x  = long, y = lat , 
                   group=zone, 
                   fill = mean_rupture,
                   alpha=reporting_rate)) +
  theme_minimal() +
  coord_map()+ 
  scale_fill_gradient(name = "Mean reported stockout days",
                      high = '#ef0404', low = '#6ddb3f',
                       space = "Lab",
                      na.value = "grey50", guide = "colourbar")#+ 

#  theme(legend.position = "bottom")




```

## Heatmap

```{r, fig.height=500, fig.width = 50 , limitsize = FALSE}
library(zoo)

org_u <- snis_org_units[,!(names(snis_org_units) %in% c('date_opening', 'coordinates'))]
colnames(org_u) <- c('zone', 'zone_name', 'region')
data_rupture2 <- merge(data_rupture, org_u, by = 'zone')

data_rupture2$month <- as.factor(as.yearmon(paste0(as.character(data_rupture2$period), '15'), format="%Y%m%d"))
data_test <- data_rupture2[data_rupture2$zone %in% unique(data_rupture$zone),]

unique_fosa <- unique(data_test[,c('zone', "org_unit_ID" )])
ord_fosa <- unique_fosa$org_unit_ID[order(unique_fosa$zone)]
data_test$org_unit_ID = factor(data_test$org_unit_ID, ord_fosa)

unique_zone <- unique(data_test[,c('zone', "region" )])
ord_reg <- unique_zone$zone[order(unique_zone$region)]
data_test$zone = factor(data_test$zone, ord_reg)

d <- ggplot(data_test, aes(month, org_unit_ID, fill = value))
d + geom_bin2d( ) + 
  scale_fill_gradient(#name = "Mean reported stockout days",
                      high = '#ef0404', low = '#6ddb3f',
                       space = "Lab",
                      na.value = "grey50", guide = FALSE) +
  facet_grid(region + zone ~ ., space = "free" , scales = 'free')+
  theme( axis.text.y=element_blank(),
         axis.ticks.y=element_blank())



#+ scale_x_date(date_breaks = "1 month", date_labels = "%m")+ 
  
  
ggsave('test.pdf', device = 'pdf', limitsize = FALSE)

```

```{r}
rupture_indic <- unique(data$name)[grep( 'rupture', unique(data$name))]

for(indic_name in rupture_indic){
  data_rupture <- data[data$name == indic_name & 
                            data$value < 32 ,]
  stockout_by_zone <- group_by(data_rupture , zone)
  mean_reported_stockout <- summarize(stockout_by_zone, mean_rupture = mean(value))
  map_reported_stockout <- right_join(coordinates , mean_reported_stockout, by = 'zone')
  map_reported_stockout_2 <- merge(map_reported_stockout, map_reporting_rates,
                                   by=c('long','lat','order','hole','piece','zone','group') , sort = FALSE)
  
  title <- gsub(' - jours rupture de stock', '', indic_name)
a <- ggplot(map_reported_stockout_2)+
  geom_polygon(aes(x  = long, y = lat , 
                   group=zone, 
                   fill = mean_rupture,
                   alpha=reporting_rate)) +
  theme_minimal() + ggtitle(title) +
  coord_map()+ 
  scale_fill_gradient(name = "Mean reported stockout days",
                      high = '#ef0404', low = '#6ddb3f',
                       space = "Lab",
                      na.value = "grey50", guide = "colourbar")

  print(a)

  
}

```

```{r}
ff <- merge(org_u, mean_reported_stockout, by.x = 'zone' , by.y = 'zone')

library(plyr)
gg <- ddply(ff, .(region) , function(x) mean(x$mean_rupture))

ggplot(ff , aes(mean_rupture)) +
  geom_density() +
  facet_wrap(~region)





hist(gg$V1)
```

1. add region
2. dist of rupture by region
3. funnel plot


## PDSS data

```{r}

pdss_bcg <- data_pdss[as.character(data_pdss$name) %in% c('13.5.1 BCG, Penta, la poliomyélite, la fièvre jaune, hépatite B, la rougeole, le tétanos', '13.5.1 BCG'), ]

#'13.5.1 BCG',

pdss_bcg$stockout <- pdss_bcg$value == 0
pdss_bcg <- merge(pdss_bcg , matched_zones, by.x = 'zone' , by.y = 'id_pdss')
pdss_bcg$id_snis <- as.character(pdss_bcg$id_snis)


pdss_bcg_zone <- group_by(pdss_bcg , id_snis)
pdss_bcg_stockout <- summarise(pdss_bcg_zone, mean_stockout = mean(stockout))

pdss_snis_compare <- merge(pdss_bcg_stockout, mean_reported_stockout, by.x = 'id_snis', by.y = 'zone')
plot(pdss_snis_compare$mean_stockout , pdss_snis_compare$mean_rupture)

```